---
output:
  html_document:
    toc: yes
    toc_depth: 6
    toc_float: yes
  github_document:
    toc: yes
    toc_depth: 6
---

<meta name="viewport" content="width=device-width, initial-scale=1">
<style>
body {
  margin: 0;
  font-family: Arial, Helvetica, sans-serif;
}

.topnav {
  overflow: hidden;
  background-color: #333;
}

.topnav a {
  float: left;
  color: #f2f2f2;
  text-align: center;
  padding: 14px 16px;
  text-decoration: none;
  font-size: 17px;
}

.topnav a:hover {
  background-color: #ddd;
  color: black;
}

.topnav a.active {
  background-color: #4CAF50;
  color: white;
}
</style>

<div class="topnav">
  <a class="active" href="./index">Home</a>
  <a href="./data">Data</a>
  <a href="./EDA.html">Exploratory Data Analysis</a>
  <a href="./gdp_modeling.html">Modeling of GDP</a>
  <a href="https://github.com/sjmiller8182/gdp-pred/blob/master/analysis/EDA.Rmd">GitHub</a>
</div>




# Exploratory Data Analysis of Economic Indicators

**Stuart Miller**  
**`r Sys.time()`**

```{r setup, include=FALSE, message=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(tswge)
library(GGally)
library(gridExtra)
source('../helper/plotting.R')
```

```{r, read-data, include=FALSE}
#read data with set columns
data <- read_csv('../data/economic_indicators_all_ex_3mo_china.csv',
                 col_types = cols(
                   date = col_character(),
                   gdp_change = col_double(),
                   unrate = col_double(),
                   nfjobs = col_double(),
                   treas10yr = col_double(),
                   fedintrate = col_double(),
                   personincomechg = col_double(),
                   corpprofitchg = col_double()
                 )
              )
```



## Multivariate

Look at correlation between gdp and lags of the other variables.

### GDP Change vs Unemployment Rate

There is evidence of a weak correlation between `gdp_change` and the first 10 lags of `unrate`.

```{r, fig.width=10, fig.height=10}
plot.cross.lags(data, 'gdp_change', 'unrate', 30)
```




### GDP Change vs Non-farm Jobs

There is evidence of a weak correlation between `gdp_change` and the first 10 lags of `nfjobs`.

```{r, fig.width=10, fig.height=10}
plot.cross.lags(data, 'gdp_change', 'nfjobs', 20)
```


### GDP Change vs 10 Year Treasury Yield

There is evidence of a weak correlation between `gdp_change` and the first 10 lags of `treas10yr`.

```{r, fig.width=10, fig.height=10}
plot.cross.lags(data, 'gdp_change', 'treas10yr', 20)
```







### GDP Change vs Federal Interest Rate

There is evidence of a weak correlation between `gdp_change` and the first 10 lags of `fedintrate`.

```{r, fig.width=10, fig.height=10}
plot.cross.lags(data, 'gdp_change', 'fedintrate', 19)
```




### GDP Change vs Change in Personal Income

There is evidence of a weak correlation between `gdp_change` and the first 4 lags of `personincomechg`.

```{r, fig.width=10, fig.height=10}
plot.cross.lags(data, 'gdp_change', 'personincomechg', 20)
```


### GDP Change vs Change in Corparate Profits

There does not seem to be a correlation between change in GDP and lags of change in corparate profits.

```{r, fig.width=10, fig.height=10}
plot.cross.lags(data, 'gdp_change', 'corpprofitchg', 19)
```

### GDP Change vs CPI



```{r, fig.width=10, fig.height=10}
plot.cross.lags(data, 'gdp_change', 'cpi', 20)
```


### GDP Change vs population



```{r, fig.width=10, fig.height=10}
plot.cross.lags(data, 'gdp_change', 'population', 20)
```


### GDP Change vs crude_wti



```{r, fig.width=10, fig.height=10}
plot.cross.lags(data, 'gdp_change', 'crude_wti', 20)
```



### GDP Change vs ppi



```{r, fig.width=10, fig.height=10}
plot.cross.lags(data, 'gdp_change', 'ppi', 20)
```


### GDP Change vs gold



```{r, fig.width=10, fig.height=10}
plot.cross.lags(data, 'gdp_change', 'gold', 20)
```

### GDP Change vs japan



```{r, fig.width=10, fig.height=10}
plot.cross.lags(data, 'gdp_change', 'japan', 20)
```


### GDP Change vs UK



```{r, fig.width=10, fig.height=10}
plot.cross.lags(data, 'gdp_change', 'uk', 20)
```



```{r, fig.width=10, fig.height=10}
ggpairs(data %>% select(-c(date, gdp_change)))
```





